\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2019-02-07 document/final/XSGRB_sample_final.tex
Accepted version

\title{The X-shooter GRB afterglow legacy sample (XS-GRB)\thanks{Based on
		observations collected at the European Southern Observatory, Paranal, Chile,
		Program IDs: 084.A-0260, 084.D-0265, 085.A-0009, 086.A-0073, 087.A-0055,
		088.A-0051, 089.A-0067, 090.A-0088, 091.A-0877, 0091.C-0934, 092.A-0124,
		092.D-0056, 092.D-0633, 093.A-0069, 094.A-0134, 095.A-0045, 095.B-0811,
		096.A-0079, 097.A-0036, 098.A-0136, and 098.A-0055.}}

@@

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph
	until \termdate. In total, we have obtained spectroscopic observations of 103 individual GRBs
	observed within 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimize biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneously selected sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of
	the underlying host galaxies. We provide an assessment of the degree of
	completeness with respect to the parent GRB population, in terms of the X-ray
	properties of the bursts in the sample and find that the sample presented here
	is representative of the full \textit{Swift} sample. We constrain the fraction
	of dark bursts to be < 28 per cent and confirm previous results that higher
	optical darkness is correlated with increased X-ray absorption. For the 42
	bursts for which it is possible, we provide a measurement of the neutral
	hydrogen column density, increasing the total number of published H{\sc I}
	column density measurements by $\sim$ 33 per cent. This dataset provides a
	unique resource to study the ISM across cosmic time, from the local progenitor
	surroundings to the intervening Universe.}

@@

Gamma-ray bursts (GRBs) are bright glimpses of electromagnetic radiation that
pierce through the universe, all the way from the edges of the observable
universe. They provide constraints on a very wide range of topics in
astrophysics. Examples range from small-scale phenomena relating to magnetars,
properties of highly relativistic jets, hyper/supernova explosions, the
interstellar medium, dust extinction curves, starburst galaxies, chemical and
molecular abundances, escape of ionizing radiation, the ionization state of the
intergalactic medium, intervening absorption systems to standard candles in
cosmology \citep[e.g.,][]{Wijers1998, Savaglio2006, Ghirlanda2007, Molinari2007,
	Amati2008, Vergani2009, Prochaska2009, HjorthBloom2012, Rowlinson2017,
	Christensen2017}.

The \textit{Neil Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of the
intergalactic medium, which was already hinted at by results from earlier
missions \citep[e.g.,][]{Paradijs2000, Ricker2004}. With the three on-board
instruments, the Burst Alert Telescope \citep[BAT;][]{Barthelmy2005}, the X-Ray
Telescope \citep[XRT;][]{Burrows2005}, and the UltraViolet and Optical Telescope
\citep[UVOT;][]{Roming2005}, \swift~is an ideal observatory for GRB hunting. A
crucial aspect of the success of the {\it Swift} mission has been the extensive
ground-based follow-up observations of the afterglows and of the host galaxies
of the GRBs, involving a large community of researchers. This fruitful
collaboration has been facilitated by the open data access policy of the {\it
	Swift} mission. The close collaboration between detection facilities and
electromagnetic  follow-up campaigns continue to be immensely rewarding, as
recently highlighted by the simultaneous detection of gravitational waves and
light from the neutron star merger in the shape of GW170817/GRB170817A/AT2017gfo
\citep{2017ApJ...848L..14G, 2017ApJ...848L..15S, 2017ApJ...848L..16S,
	2017ApJ...848L..17C, 2017ApJ...848L..18N, 2017ApJ...848L..19C,
	2017ApJ...848L..27T, 2017ApJ...850L...1L, 2017Natur.551...64A,
	2017Natur.551...67P, 2017Natur.551...71T, 2017Natur.551...75S,
	2017Sci...358.1583K, 2017Sci...358.1556C, Burgess2017, LIGOScientificCollaboration2017a,
	LIGOScientificCollaboration2017}.

In the beginning of the {\it Swift} era most of the follow-up afterglow
spectroscopy was secured using low-resolution spectrographs \citep[typically
$R=\lambda/\Delta\lambda$$<$1000, e.g.,][]{Fynbo2009}. Spectroscopy is powerful
as it allows us to secure information even for very faint targets
\citep{Kruhler2012}. This allows the measurement of a number of important
parameters such as redshifts, spectral slopes, and extinction. For a handful of
very bright afterglows high-resolution (typically R$>$20000) spectra have been
secured, and for these events much more information about conditions inside the
host galaxies were extracted \citep[e.g.,][]{Fiore2005, Thone2007,
	Prochaska2007, Vreeswijk2007, DElia2009, Castro-Tirado2010}.

@@

In this paper, we present the results of a dedicated effort over the years 2009
-- 2017 to use the X-shooter spectrograph to secure spectroscopic observations
of afterglows and host galaxies of GRBs detected by {\it Swift}. We here make
all the data resulting from the survey publicly available in reduced form (see Sect
\ref{products}).

The paper is organized in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy. In Sect.~\ref{obs}, we describe the observations and the instrumental
setups, and in Sect.~\ref{proc} we detail the methodological strategies adopted
in the data reduction process and auxiliary material. In Sect.~\ref{results} we
describe the results of the survey, i.e. the efficiency of the follow-up effort
and the characteristics of the observed bursts. We also assess the
completeness of the realized sample. Finally, we offer our conclusions
in Sect.~\ref{conclusions}. We use the $\Lambda$CDM cosmology parameters
provided by the \citet{Planck2015} in which the universe is flat with $H_0 =
67.7$\,km\,s$^{-1}$ Mpc$^{-1}$ and $\Omega_\mathrm{m} = 0.307$.

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs without hampering the follow-up effort. Many natural follow-up
restrictions exist already, being it weather conditions, pointing restrictions
of the telescope or unconstrained burst localizations as reported by the alerting
facility. To maximize the return of the follow-up campaign, we have chosen a
few selection criteria that facilitate an unbiased selection of bursts, while
at the same time allowing for a high follow-up success rate. The importance of
defining unbiased selection criteria has been highlighted previously
\citep{Jakobsson2006b, Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a},
when trying to address the intrinsic underlying distribution functions such as
the redshift distribution, host metallicity distribution, or afterglow
brightness distribution. When investigating a specific distribution function, a
high degree of \textit{completeness} is desired \citep[e.g.,][]{Perley2016b}.

In defining the selection criteria, we simultaneously aim to minimize any biases
against intrinsic astrophysical conditions while at the same time maximizing the
likelihood of successful observations, hence allowing us to obtain a higher
degree of completeness. By restricting the selection criteria to conditions
local to the Milky Way and therefore independent of intrinsic GRB properties,
the aim is that the collected sample represents the underlying distribution of
GRBs in a fair way. The selection criteria used here are based on previous,
similar studies \citep{Jakobsson2006b, Fynbo2009, Hjorth2012}. We characterize
the sample completeness in Sect. \ref{completeness}.

The selection criteria that define a GRB as part of our initial \textit{statistical}
	sample are:

@@

	\item XRT started observing within 10 minutes after the GRB; an XRT position
	must be distributed within 12 hr.

	\item The target must be visible from Cerro Paranal for at least 60 minutes, 30
degrees above the horizon, with the Sun below $-12$ degrees\footnote{Note that
	in the P84 proposal the criteria have been stated differently, the visibility
	constraint being replaced by a declination + Sun angle constraint. The criteria
	listed above are, however, those defining the sample.}.

@@

Because the localization accuracy of BAT is 1 -- 4 arcminutes
\citep{Barthelmy2005}, an afterglow identification based on BAT alone would be
harder, and  host association impossible. We therefore additionally require an
X-ray position from XRT to be distributed to the GCN network
\citep{Barthelmy2000} within 12 hours and to account for observing constraints
on \textit{Swift} that XRT began observations within 10 minutes. The additional
timing requirement of the XRT follow-up means that all bursts in our sample have
detected X-ray afterglows. Because the XRT completeness is very high for
promptly-repointed GRBs \citep{Burrows2007}, this cut should not alter the
parent sample significantly.

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to be completed. Since the GRB population is isotropically
distributed on the sky as seen from Earth, and because the GRB properties do not
depend on position on the sky \citep{Meegan1992, Briggs1996, Ukwatta2016}, this
cut does not influence our ability to fairly sample the underlying GRB
population. The same arguments apply to the requirement that there are no nearby
foreground bright stars. We additionally require that the Galactic extinction is
below $A_V \lesssim 0.5$ mag, based on the extinction maps by
\citet{Schlegel1998}\footnote{We use the updated values of \citet{Schlafly2011}
	to correct for foreground extinction, but the sample criterion is based on the
	old \citet{Schlegel1998} values for consistency with the first semesters of our
	program.}. Choosing low-extinction sightlines also reduces the problem of field
crowding and the contamination from Galactic outbursts posing as GRB impostors.
These additional cuts should not influence the optical properties of the bursts
themselves, only our ability to successfully secure the observations that allow
us to investigate the spectroscopic properties of GRBs.

Bursts that fulfill these selection criteria are what we define as our initial
\textit{statistical} sample. To reach a higher degree of sample completeness, we
impose additional cuts that increase our ability to infer population properties.
These are presented in Sect. \ref{badbursts}. Using this sample we will be able
to address population properties of \swift-detected bursts. We further discuss
the effect of these selection criteria and their implication for the
completeness of the sample in Sect. \ref{completeness}.

@@

Our collaboration has set up a procedure to promptly react to alerts from
\swift~and other satellites (such as \textit{Fermi}). We process automatically
GCN notices in order to flag those events belonging to our statistical sample.
Two people are permanently on alert to manually supervise each event, for
example to recognize events of special interest beyond their inclusion in the
sample. Members of our collaboration have access to numerous facilities spread
throughout the world, among which the William Herschel Telescope (WHT), Nordic
Optical Telescope (NOT), the Gran Telescopio Canarias (GTC), the Telescopio
Nazionale Galileo (TNG), the Gamma-Ray Burst Optical Near-infrared Detector
(GROND \citealt{Greiner2008}), and the Xinlong observatory. Thanks to this
network, we can often conduct searches for optical/NIR afterglows ahead of the
time of the X-shooter spectroscopy, thus helping to plan the spectroscopic
observations. If this is not possible, we use the built-in acquisition camera of
X-shooter to search for counterparts, and adapt the observing strategy on the
fly. In many cases we could interact directly with the staff at the telescope to
aid the observation. Raw data are usually available within minutes in the
\href{http://archive.eso.org/wdb/wdb/eso/xshooter/form}{ESO archive} and are
promptly reduced by members of our collaboration, often using archival
calibration data which are readily available. This allows us in most cases to
report the preliminary results (redshift, identification of the most prominent
emission and absorption feature, etc.) within a few hours after the beginning of
the observations and plan additional X-shooter observations if deemed necessary.

@@

\begin{enumerate}
	\item The GRB triggered \textit{Swift}.
	\item The X-ray position must be available less than 1 hr after the GRB.
	%    \item X-shooter is the currently operating instrument at the telescope.
	% DM I removed the line above because we do send the trigger even if X-shooter is not mounted (though, the trigger is rejected, as explained in the text above)
	%    \item GRB promptly visible from Cerro Paranal.
	%    \item Error radius < 60\arcsec.
	%    \item No limit on declination.
	%    \item No constraint on Moon phase or distance.
	\item The elevation of the source in the sky is $> 22^\circ$ (both at trigger time and 15 min after).
	\item The Sun elevation from Paranal is $< -12^\circ$ (both at trigger time and 15 min after).
	\item Galactic $A_V \leq 1.0$~mag according to the maps of \citet{Schlegel1998}.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}.
\end{enumerate}

@@

The use of RRM is unique as it allows to sample a long logarithmic time span in
the GRB lifetime, and it exploits the extreme brightness of early afterglows. A
pivotal example is the study of temporal variability of GRB afterglow absorption
systems due to effect of the GRB itself on the surrounding medium \citep[e.g.,
see][]{Dessauges-Zavadsky2006, Vreeswijk2007, DElia2009, Vreeswijk2013}.

@@

\begin{figure*}
	\centerline{\includegraphics[width=\linewidth]{figures/skymap.pdf}} \caption{Mollweide projection
	in Galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. The empty
	stars/circles are the positions of all the 1266 \textit{Swift} bursts detected
	until \termdate. Stars indicate bursts with measured redshifts and circles
	indicate those without. Blue stars indicate the position of the 104 bursts
	fulfilling the sample criteria specified in Sect. \ref{samplecrit} that have a
	measured redshift. Red outlines are added to the 84 GRBs that enter our
	statistical sample with both X-shooter spectroscopy and a measured redshift.
	Red outlines of empty stars represent bursts which have been observed with
	X-shooter and has a measured redshift, but is not a part of our statistical
	sample. The blue dots show the positions of the 61 GRBs of our statistical
	sample that lack redshift measurements. A red outline is added around the six
	bursts in our statistical sample which were observed with X-shooter, but did
	not yield a redshift measurement. Two empty circles with red outlines indicate
	bursts outside the statistical sample that were followed up with X-shooter, but
	without a redshift measurement. The different samples are compared in Sect.
	\ref{results}. The background shows the dust maps presented in
	\citet{Schlegel1998}. Please note that we removed the background where the
	sample criterion is violated ($A_V > 0.5$ mag) and replaced it with a white
	background. The grey scale bar below indicates the value of $A_V$ on the plot.
	The dotted lines indicate intervals of 30$^\circ$ in longitude and latitude.}
\label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during the
duration of this follow-up campaign. The observations have been taken during a
period of eight years corresponding to the ESO observing periods P84 through P98
under the following programme IDs: 084.A-0260, 085.A-0009, 086.A-0073,
087.A-0055, 088.A-0051, 089.A-0067, 090.A-0088, 092.A-0124, 093.A-0069,
094.A-0134, 095.A-0045, 096.A-0079, 097.A-0036, and 098.A-0055 (PI: Fynbo) and
0091.C-0934 (PI: Kaper). These proposals were initiated on Guaranteed Time. We
have included a few additional bursts, from the programmes 084.D-0265 (PI:
Benetti), 091.A-0877 (PI: Schady), 092.D-0056 (PI: Rau), 092.D-0633,
098.A-0136 (PI: Greiner), and 095.B-0811 (PI: Levan). The total collection of
spectra represents \textit{all} GRB afterglows that have been followed up by
X-shooter up to \termdate, which marks the end of the XS-GRB legacy follow-up
program.

@@

X-shooter covers the spectral wavelength region from 300 nm to 2480 nm in a
single exposure, by splitting the light into three separate spectroscopic arms
through the use of two dichroics. The ultraviolet blue (UVB) arm covers 300 -
550 nm, the visual (VIS) arm covers 550 - 1020 nm, and the near-infrared (NIR)
arm covers 1020 - 2480 nm. For some of the observations, we have applied a
$K$-band blocking filter, cutting the coverage of the NIR arm at 2100 nm. The
$K$-band blocking filter is only used after 2012, where it was installed. This
is done to reduce the amount of scattered background light from the thermal
infrared. For the majority of observations, a nodding observing scheme
has been employed, with a nodding throw of 5\arcsec. Each nodding observation
has typically been carried out in a standard ABBA pattern. For some cases,
conditions during the follow-up (either technical or weather), have necessitated
alterations to this scheme as described in App. \ref{notes}. For RRM triggers, a
slightly different observing strategy was employed. Starting as rapidly as
possible, a simple stare mode sequence was started, with 5 spectroscopic
integrations with increasing exposure times.

@@

Due to a mechanical failure, the atmospheric dispersion corrector (ADC) was
disabled from 1st of August 2012 until the end of this program. Only GRB~100728B
was affected by the failing ADC prior to disablement, resulting in a
lower-than-nominal throughput. To avoid chromatic slit-loss due to atmospheric
dispersion, nearly all subsequent observations have been carried out at
parallactic angle. A consequence of this is that for all observations following
1st of August 2012, the centroid of the trace of the source changes position
across the spatial direction of the slit as a function of wavelength. This
effect has been modeled in the extraction procedure, as described in Sect.
\ref{extract}.

@@

For all Earth-based telescopes, the light first has to pass through Earth's
atmosphere, where the atmospheric content and conditions make an imprint on the
received spectrum. These telluric features can be corrected for in a multitude
of ways. We employ a prioritized list of methods here, depending on the
availability of the chosen methods. Since the observations are often taken at
odd times under varying conditions, this prioritized list ensures that we are
always doing the best possible correction.

>>>>

\title{The X-shooter GRB afterglow legacy sample (XS-GRB)\thanks{Based on
		observations collected at the European Southern Observatory, Paranal, Chile,
		Programme IDs: 084.A-0260, 084.D-0265, 085.A-0009, 086.A-0073, 087.A-0055,
		088.A-0051, 089.A-0067, 090.A-0088, 091.A-0877, 0091.C-0934, 092.A-0124,
		092.D-0056, 092.D-0633, 093.A-0069, 094.A-0134, 095.A-0045, 095.B-0811,
		096.A-0079, 097.A-0036, 098.A-0136, and 098.A-0055.}}

@@

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph
	until \termdate. In total, we have obtained spectroscopic observations of 103 individual GRBs
	observed within 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimise biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneously selected sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of
	the underlying host galaxies. We provide an assessment of the degree of
	completeness with respect to the parent GRB population, in terms of the X-ray
	properties of the bursts in the sample and find that the sample presented here
	is representative of the full \textit{Swift} sample. We have constrained the fraction
	of dark bursts to be < 28 per cent and confirm previous results that higher
	optical darkness is correlated with increased X-ray absorption. For the 42
	bursts for which it is possible, we have provided a measurement of the neutral
	hydrogen column density, increasing the total number of published H{\sc I}
	column density measurements by $\sim$ 33 per cent. This dataset provides a
	unique resource to study the ISM across cosmic time, from the local progenitor
	surroundings to the intervening Universe.}

@@

Gamma-ray bursts (GRBs) are bright glimpses of electromagnetic radiation that
pierce through the universe, all the way from the edges of the observable
universe. They provide constraints on a very wide range of topics in
astrophysics. Examples range from small-scale phenomena relating to magnetars,
properties of highly relativistic jets, hypernova and supernova explosions, the
interstellar medium, dust extinction curves, starburst galaxies, chemical and
molecular abundances, escape of ionising radiation, the ionization state of the
intergalactic medium, intervening absorption systems to standard candles in
cosmology \citep[e.g.][]{Wijers1998, Savaglio2006, Ghirlanda2007, Molinari2007,
	Amati2008, Vergani2009, Prochaska2009, HjorthBloom2012, Rowlinson2017,
	Christensen2017}.

The \textit{Neil Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of the
intergalactic medium, which has already been hinted at by results from earlier
missions \citep[e.g.][]{Paradijs2000, Ricker2004}. With the three on-board
instruments, the Burst Alert Telescope \citep[BAT;][]{Barthelmy2005}, the X-Ray
Telescope \citep[XRT;][]{Burrows2005}, and the UltraViolet and Optical Telescope
\citep[UVOT;][]{Roming2005}, \swift~is an ideal observatory for GRB hunting. A
crucial aspect of the success of the {\it Swift} mission has been the extensive
ground-based follow-up observations of the afterglows and of the host galaxies
of the GRBs, involving a large community of researchers. This fruitful
collaboration has been facilitated by the open data access policy of the {\it
	Swift} mission. The close collaboration between detection facilities and
electromagnetic  follow-up campaigns continue to be immensely rewarding, as
recently highlighted by the simultaneous detection of gravitational waves and
light from the neutron star merger in the shape of GW170817/GRB170817A/AT2017gfo
\citep{2017ApJ...848L..14G, 2017ApJ...848L..15S, 2017ApJ...848L..16S,
	2017ApJ...848L..17C, 2017ApJ...848L..18N, 2017ApJ...848L..19C,
	2017ApJ...848L..27T, 2017ApJ...850L...1L, 2017Natur.551...64A,
	2017Natur.551...67P, 2017Natur.551...71T, 2017Natur.551...75S,
	2017Sci...358.1583K, 2017Sci...358.1556C, Burgess2017, LIGOScientificCollaboration2017a,
	LIGOScientificCollaboration2017}.

In the beginning of the {\it Swift} era most of the follow-up afterglow
spectroscopy was secured using low-resolution spectrographs \citep[typically
$R=\lambda/\Delta\lambda$$<$1000, e.g.][]{Fynbo2009}. Spectroscopy is powerful
as it allows us to secure information even for very faint targets
\citep{Kruhler2012}. This allows the measurement of a number of important
parameters such as redshifts, spectral slopes, and extinction. For a handful of
very bright afterglows high-resolution (typically R$>$20000) spectra have been
secured, and for these events much more information about conditions inside the
host galaxies were extracted \citep[e.g.][]{Fiore2005, Thone2007,
	Prochaska2007, Vreeswijk2007, DElia2009, Castro-Tirado2010}.

@@

In this paper, we present the results of a dedicated effort over the years 2009
-- 2017 to use the X-shooter spectrograph to secure spectroscopic observations
of afterglows and host galaxies of GRBs detected by {\it Swift}. Here we make
all the data resulting from the survey publicly available in reduced form (see Sect
\ref{products}).

The paper is organised in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy. In Sect.~\ref{obs}, we describe the observations and the instrumental
setups, and in Sect.~\ref{proc} we detail the methodological strategies adopted
in the data reduction process and auxiliary material. In Sect.~\ref{results} we
describe the results of the survey, the efficiency of the follow-up effort,
and the characteristics of the observed bursts. We also assess the
completeness of the realised sample. Finally, we offer our conclusions
in Sect.~\ref{conclusions}. We use the $\Lambda$CDM cosmology parameters
provided by the \citet{Planck2015} in which the universe is flat with $H_0 =
67.7$\,km\,s$^{-1}$ Mpc$^{-1}$ and $\Omega_\mathrm{m} = 0.307$.

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs without hampering the follow-up effort. Many natural follow-up
restrictions exist already, being it weather conditions, pointing restrictions
of the telescope or unconstrained burst localizations as reported by the alerting
facility. To maximise the return of the follow-up campaign, we have chosen a
few selection criteria that facilitate an unbiased selection of bursts, while
at the same time allowing for a high follow-up success rate. The importance of
defining unbiased selection criteria has been highlighted previously
\citep{Jakobsson2006b, Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a},
when trying to address the intrinsic underlying distribution functions such as
the redshift distribution, host metallicity distribution, or afterglow
brightness distribution. When investigating a specific distribution function, a
high degree of completeness is desired \citep[e.g.][]{Perley2016b}.

In defining the selection criteria, we simultaneously aim to minimise any biases
against intrinsic astrophysical conditions while at the same time maximising the
likelihood of successful observations, hence allowing us to obtain a higher
degree of completeness. By restricting the selection criteria to conditions
local to the Milky Way and therefore independent of intrinsic GRB properties,
the aim is that the collected sample represents the underlying distribution of
GRBs in a fair way. The selection criteria used here are based on previous,
similar studies \citep{Jakobsson2006b, Fynbo2009, Hjorth2012}. We characterise
the sample completeness in Sect. \ref{completeness}.

The selection criteria that define a GRB as part of our initial statistical
	sample are:

@@

	\item XRT started observing within ten minutes after the GRB; an XRT position
	must be distributed within 12 hr.

	\item The target must be visible from Cerro Paranal for at least 60 minutes, 30
degrees above the horizon, with the Sun below $-12$ degrees\footnote{We note that
	in the P84 proposal the criteria have been stated differently, the visibility
	constraint being replaced by a declination + Sun angle constraint. The criteria
	listed above are, however, those defining the sample.}.

@@

Because the localization accuracy of BAT is 1 -- 4 arcminutes
\citep{Barthelmy2005}, an afterglow identification based on BAT alone would be
harder, and  host association impossible. We therefore additionally require an
X-ray position from XRT to be distributed to the GCN network
\citep{Barthelmy2000} within 12 hours and to account for observing constraints
on \textit{Swift} that XRT began observations within ten minutes. The additional
timing requirement of the XRT follow-up means that all bursts in our sample have
detected X-ray afterglows. Because the XRT completeness is very high for
promptly-repointed GRBs \citep{Burrows2007}, this cut should not alter the
parent sample significantly.

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least one hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to be completed. Since the GRB population is isotropically
distributed on the sky as seen from Earth, and because the GRB properties do not
depend on position on the sky \citep{Meegan1992, Briggs1996, Ukwatta2016}, this
cut does not influence our ability to fairly sample the underlying GRB
population. The same arguments apply to the requirement that there are no nearby
foreground bright stars. We additionally require that the Galactic extinction is
below $A_V \lesssim 0.5$ mag, based on the extinction maps by
\citet{Schlegel1998}\footnote{We use the updated values of \citet{Schlafly2011}
	to correct for foreground extinction, but the sample criterion is based on the
	old \citet{Schlegel1998} values for consistency with the first semesters of our
	programme.}. Choosing low-extinction sightlines also reduces the problem of field
crowding and the contamination from Galactic outbursts posing as GRB impostors.
These additional cuts should not influence the optical properties of the bursts
themselves, only our ability to successfully secure the observations that allow
us to investigate the spectroscopic properties of GRBs.

Bursts that fulfill these selection criteria are what we define as our initial
statistical sample. To reach a higher degree of sample completeness, we
impose additional cuts that increase our ability to infer population properties.
These are presented in Sect. \ref{badbursts}. Using this sample we will be able
to address population properties of \swift-detected bursts. We further discuss
the effect of these selection criteria and their implication for the
completeness of the sample in Sect. \ref{completeness}.

@@

Our collaboration has set up a procedure to promptly react to alerts from
\swift~and other satellites (such as \textit{Fermi}). We process automatically
GCN notices in order to flag those events belonging to our statistical sample.
Two people are permanently on alert to manually supervise each event, for
example to recognise events of special interest beyond their inclusion in the
sample. Members of our collaboration have access to numerous facilities spread
throughout the world, among which the William Herschel Telescope (WHT), Nordic
Optical Telescope (NOT), the Gran Telescopio Canarias (GTC), the Telescopio
Nazionale Galileo (TNG), the Gamma-Ray Burst Optical Near-infrared Detector
(GROND \citealt{Greiner2008}), and the Xinlong observatory. Thanks to this
network, we can often conduct searches for optical/NIR afterglows ahead of the
time of the X-shooter spectroscopy, thus helping to plan the spectroscopic
observations. If this is not possible, we use the built-in acquisition camera of
X-shooter to search for counterparts, and adapt the observing strategy on the
fly. In many cases we could interact directly with the staff at the telescope to
aid the observation. Raw data are usually available within minutes in the
\href{http://archive.eso.org/wdb/wdb/eso/xshooter/form}{ESO archive} and are
promptly reduced by members of our collaboration, often using archival
calibration data which are readily available. This allows us in most cases to
report the preliminary results (redshift, identification of the most prominent
emission and absorption feature, etc.) within a few hours after the beginning of
the observations and plan additional X-shooter observations if deemed necessary.

@@

\begin{enumerate}
	\item The GRB triggered \textit{Swift}.
	\item The X-ray position must be available less than one hour after the GRB.
	%    \item X-shooter is the currently operating instrument at the telescope.
	% DM I removed the line above because we do send the trigger even if X-shooter is not mounted (though, the trigger is rejected, as explained in the text above)
	%    \item GRB promptly visible from Cerro Paranal.
	%    \item Error radius < 60\arcsec.
	%    \item No limit on declination.
	%    \item No constraint on Moon phase or distance.
	\item The elevation of the source in the sky is $> 22^\circ$ (both at trigger time and 15 min after).
	\item The Sun elevation from Paranal is $< -12^\circ$ (both at trigger time and 15 min after).
	\item Galactic $A_V \leq 1.0$~mag according to the maps of \citet{Schlegel1998}.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}.
\end{enumerate}

@@

The use of RRM is unique as it allows to sample a long logarithmic time span in
the GRB lifetime, and it exploits the extreme brightness of early afterglows. A
pivotal example is the study of temporal variability of GRB afterglow absorption
systems due to effect of the GRB itself on the surrounding medium \citep[e.g.
see][]{Dessauges-Zavadsky2006, Vreeswijk2007, DElia2009, Vreeswijk2013}.

@@

\begin{figure*}
	\centerline{\includegraphics[width=\linewidth]{figures/skymap.pdf}} \caption{Mollweide projection
	in Galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. The empty
	stars/circles are the positions of all the 1266 \textit{Swift} bursts detected
	until \termdate. Stars indicate bursts with measured redshifts and circles
	indicate those without. Blue stars indicate the position of the 104 bursts
	fulfilling the sample criteria specified in Sect. \ref{samplecrit} that have a
	measured redshift. Red outlines are added to the 84 GRBs that enter our
	statistical sample with both X-shooter spectroscopy and a measured redshift.
	Red outlines of empty stars represent bursts which have been observed with
	X-shooter and has a measured redshift, but is not a part of our statistical
	sample. The blue dots show the positions of the 61 GRBs of our statistical
	sample that lack redshift measurements. A red outline is added around the six
	bursts in our statistical sample which were observed with X-shooter, but did
	not yield a redshift measurement. Two empty circles with red outlines indicate
	bursts outside the statistical sample that were followed up with X-shooter, but
	without a redshift measurement. The different samples are compared in Sect.
	\ref{results}. The background shows the dust maps presented in
	\citet{Schlegel1998}. We note that we removed the background where the
	sample criterion is violated ($A_V > 0.5$ mag) and replaced it with a white
	background. The grey scale bar below indicates the value of $A_V$ on the plot.
	The dotted lines indicate intervals of 30$^\circ$ in longitude and latitude.}
\label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during the
duration of this follow-up campaign. The observations have been taken during a
period of eight years corresponding to the ESO observing periods P84 through P98
under the following programme IDs: 084.A-0260, 085.A-0009, 086.A-0073,
087.A-0055, 088.A-0051, 089.A-0067, 090.A-0088, 092.A-0124, 093.A-0069,
094.A-0134, 095.A-0045, 096.A-0079, 097.A-0036, and 098.A-0055 (PI: Fynbo) and
0091.C-0934 (PI: Kaper). These proposals were initiated on Guaranteed Time. We
have included a few additional bursts, from the programmes 084.D-0265 (PI:
Benetti), 091.A-0877 (PI: Schady), 092.D-0056 (PI: Rau), 092.D-0633,
098.A-0136 (PI: Greiner), and 095.B-0811 (PI: Levan). The total collection of
spectra represents all GRB afterglows that have been followed up by
X-shooter up to \termdate, which marks the end of the XS-GRB legacy follow-up
programme.

@@

X-shooter covers the spectral wavelength region from 300 nm to 2480 nm in a
single exposure, by splitting the light into three separate spectroscopic arms
through the use of two dichroics. The ultraviolet blue (UVB) arm covers 300 -
550 nm, the visual (VIS) arm covers 550 - 1020 nm, and the near-infrared (NIR)
arm covers 1020 - 2480 nm. For some of the observations, we have applied a
$K$-band blocking filter, cutting the coverage of the NIR arm at 2100 nm. The
$K$-band blocking filter is only used after 2012, where it was installed. This
is done to reduce the amount of scattered background light from the thermal
infrared. For the majority of observations, a nodding observing scheme
has been employed, with a nodding throw of 5\arcsec. Each nodding observation
has typically been carried out in a standard ABBA pattern. For some cases,
conditions during the follow-up (either technical or weather), have necessitated
alterations to this scheme as described in App. \ref{notes}. For RRM triggers, a
slightly different observing strategy was employed. Starting as rapidly as
possible, a simple stare mode sequence was started, with FIVE spectroscopic
integrations with increasing exposure times.

@@

Due to a mechanical failure, the atmospheric dispersion corrector (ADC) was
disabled from 1st of August 2012 until the end of this programme. Only GRB~100728B
was affected by the failing ADC prior to disablement, resulting in a
lower-than-nominal throughput. To avoid chromatic slit-loss due to atmospheric
dispersion, nearly all subsequent observations have been carried out at
parallactic angle. A consequence of this is that for all observations following
1st of August 2012, the centroid of the trace of the source changes position
across the spatial direction of the slit as a function of wavelength. This
effect has been modelled in the extraction procedure, as described in Sect.
\ref{extract}.

@@

For all Earth-based telescopes, the light first has to pass through Earth's
atmosphere, where the atmospheric content and conditions make an imprint on the
received spectrum. These telluric features can be corrected for in a multitude
of ways. We employ a prioritised list of methods here, depending on the
availability of the chosen methods. Since the observations are often taken at
odd times under varying conditions, this prioritised list ensures that we are
always doing the best possible correction.
